A Neural, Parallel Theorem Prover*

### 2019-05-13 papers/FroCoS19/paper.tex
added figure, spellcheck, clean up to 16 pages

\documentclass[runningheads]{llncs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{subcaption}
\usepackage{textcomp}
\newcommand{\lerna}{\textsc{Lerna}}
\newcommand{\z}[1]{\textsc{Z3}}
\newcommand{\mizarlarge}{\textit{M40k}}
\newcommand{\mizarsmall}{\textit{M2k}}
\begin{document}
\title{A Neural, Parallel Theorem Prover*}
\author{
	Michael Rawson*\orcidID{0000-0001-7834-1567}\inst{1} \and
	Giles Reger\inst{2}
}
\institute{
	University of Manchester, UK\\
	\email{michael@rawsons.uk}\and
	University of Manchester, UK\\
	\email{regerg@cs.man.ac.uk}
}
\maketitle

@@

\section{Introduction}
Recent advances~\cite{graph-cnn,gcn,gcn-relational} in neural network systems allow for processing graph-structured data in a neural context.
Graphs are a natural representation for logical formulae as found in automatic theorem provers (ATPs)~\cite{formula-graph}, suggesting a new breed of \emph{neural ATP} in which proof search is guided by a neural black-box acting as ``mathematician's intuition''.
Much previous work on integrating machine-learned heuristics into automatic theorem provers has relied on hand-engineered features~\cite{MaLeCoP,FEMaLeCoP,rlCoP} or other embedding methods~\cite{ENIGMA}, which have the advantage of simplicity and relative efficiency, but do not encode all information available.
By contrast, a neural method which takes into account all information should allow for greater precision in proof guidance systems.
However, in practice there are several implementation issues which must be avoided in order for neural systems to integrate with efficient traditional ATPs.
Deep Network Guided Proof Search (DNGPS)~\cite{DNGPS} is an example of previous work in this area, which integrated a deep neural guidance system into the saturation-based prover E~\cite{E}.

\begin{enumerate}
	\item Proof state in such systems may be very large, such as in saturation-based provers~\cite{Vampire}, leading to training data which is impractical to learn from and slow to evaluate.  side-stepped this by only evaluating a single clause at a time (instead of large set) and the conjecture, reducing the amount of information available for the neural network in exchange for brevity.
	\item Data structures employed may be very opaque or ``unnatural'', containing artifice designed for efficiency rather than natural comprehension by a neural network.
	\item Systems may be very sensitive to latency, which can result in the introduction of neural guidance systems crippling prover throughput and hence performance. Even with the reduced amount of data processed, DNGPS employed a two-phase approach in which the prover ran for some time with network guidance and then for the remainder without guidance.
\end{enumerate}
%
Attempting to solve these issues with a novel prover architecture, and exploring several options to improve overall efficiency, the prototype system \lerna{}\footnote{\textbf{Le}arning to \textbf{R}eason with \textbf{N}eural \textbf{A}rchitectures. Lerna is also the lair of the mythical many-headed beast \textsc{Hydra}.} takes an alternative step toward useful neural automatic theorem provers.

\section{Background}
\subsection{First-Order Logic}
(TODO copy Giles' traditional FOL-with-equality background here)

\subsection{Automatic Theorem Provers}
(TODO copy Giles' ATP background/history here)

\subsection{Machine Learning and Theorem Proving}
(TODO talk about the intersection of ATP and ML, in particular with respect to proof search guidance)

TODO The Deep Network Guided Proof Search (DNGPS) system~\cite{DNGPS} should be mentioned, as should rlCoP~\cite{rlCoP}

@@

\section{Design}
\label{section:Design}
\subsection{Rationale}
\subsection{Architecture}
\subsection{Implementation}

\section{Calculus}

@@

\subsection{Acting as a Preprocessor}
\lerna{} as a whole might also be seen as an intelligent pre-processor for existing ATPs in this setting: existing theorem provers are known to be sensitive to small changes in their input~\cite{clausification}, and generally make little attempt to split their input into smaller sub-goals, for parallelism~\cite{parallel} or otherwise.
The system can therefore act as an adaptor for any existing first-order ATP, adding parallelism opportunities and ``smoothing out'' sensitivity to input syntax.

\section{Learned Heuristic}
A suitable heuristic function for the system must predict a value between 0 and 1 for a given formula \(F\), where 0 represents a satisfiable goal and 1 represents unsatsfiability, based on a set of tagged formulae seen in previous proof search.
Although the data is collected by running the system itself (and therefore might be considered \emph{reinforcement} learning~\cite{reinforcement}), for this approach data collection and learning were considered separately and hence forms a classic supervised-learning problem.

@@

\begin{figure}
	\centering
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{translation-ast}
		\caption{AST.}
	\end{subfigure}
	\begin{subfigure}{.39\textwidth}
		\centering
		\includegraphics[width=\linewidth]{translation-graph}
		\caption{De-duplicated DAG.}
	\end{subfigure}
	\begin{subfigure}{.4\textwidth}
		\centering
		\includegraphics[width=\linewidth]{translation-nameless}
		\caption{Nameless DAG.}
	\end{subfigure}
	\begin{subfigure}{.35\textwidth}
		\centering
		\includegraphics[width=\linewidth]{translation-undirected}
		\caption{Final result.}
	\end{subfigure}
	\caption{The translation process for \(\forall X \left[p\left(f(c), X\right) \wedge \lnot\left(f(c) = d\right)\right]\) to a graph, as seen by the neural network.}
	\label{figure:translation}
\end{figure}

\subsection{Augmentation}
\label{section:augmentation}
One possible solution~\cite{imbalanced} to the problem of classification on imbalanced domains is to synthesize new data for under-represented classes --- in this case unsatisfiable formulae --- from existing data by augmenting it.
An example is augmenting image data by cropping, flipping or adding noise to existing images.

There are many possible ways to augment formulae graphs.
For this prototype, a simple approach is taken in which a small number of nonsense formulae are added to the graph by randomly adding nodes/edges where appropriate.
This approach has the advantage of exposing the network to ``noise'' such as additional axioms which might well occur in practice, but if the network is adequately capable of filtering these then no new formulae are actually seen.

\subsection{Neural Architecture}
Convolutional neural networks~\cite{cnn} are well-known as a method for processing images.
In a typical convolutional architecture, there are a series of filtering stages, followed by a densely-connected neural network.
Each filtering stage intuitively combines data from local features (via \emph{convolution}), then reduces the dimensions of the image (via \emph{pooling}) for the next stage.
Graph neural networks have analagous convolution~\cite{gcn} (combining information from neighbouring nodes) and pooling~\cite{top-k-pooling} (merging nodes to reduce the size of the input graph) operators.

A brief period of experimentation with these operators yielded the following network architecture, shown in Figure \ref{figure:network}.
It is not claimed that this is the optimal configuration, and no grid search has yet taken place to optimize the network parameters.
The network is shown processing a formula in Figure \ref{figure:computation}.
\begin{enumerate}
	\item Input. a graph \(G\) consisting of one-hot encoded nodes \(N\) and edges \(E\).
	\item Embedding. Each node is mapped to an embedding vector of size 64 via a trained dense embedding.
	\item Initial Convolution. 4 convolution layers are applied to the graph with rectified linear activations. This yields a graph of the same size, but with information exchanged between nodes.
	\item Convolution/Pooling. Similar convolution layers are then passed through \emph{top-\(k\)}~\cite{top-k-pooling} layers, retaining \(k = 60\%\) of the graph's nodes. This is repeated 3 times, reducing the size of the graph considerably.
	\item Convolution/Max-Pooling. A final convolution layer feeds into a max-pooling layer, combining all remaining node data into one datum, and dropping the edge data.
	\item Fully Connected. A fully-connected hidden layer with rectified linear activation halves the input size.
	\item Fully Connected/Softmax. A fully-connected final layer outputs two class labels, with softmax activation.
\end{enumerate}

@@

\subsection{Implementation and Training}
This architecture was implemented with the PyTorch~\cite{pytorch} neural network library, combined with a graph-processing (``geometric'') extension library, PyTorch Geometric~\cite{pytorch-geometric}, which together provide facilities for automatic differentiation, GPU-accelerated training, pre-programmed layers for graph processing, and various utilities.

The dataset is split into a large training set and a smaller test set (200 balanced examples), since unsatisfiable examples were difficult to obtain.
The training set was then augmented as described in section \ref{section:augmentation} to produce a total training set of 3.5 million examples.
The network was trained on commodity desktop hardware with a mid-range GPU~\footnote{NVIDIA\textsuperscript{\textregistered} GeForce\textsuperscript{\textregistered} GT 730.} for 8 epochs/24 hours, optimizing a negative log-likelihood loss function.

\begin{figure}
	\centering
	\captionsetup[subfigure]{justification=centering}
	\begin{subfigure}{.3\textwidth}
		\centering
		\includegraphics[width=\linewidth]{embedding}
		\caption{Output of embedding layer.}
	\end{subfigure}
	\begin{subfigure}{.3\textwidth}
		\centering
		\includegraphics[width=\linewidth]{conv}
		\caption{Output after initial convolutions.}
	\end{subfigure}
	\begin{subfigure}{.3\textwidth}
		\centering
		\includegraphics[width=\linewidth]{pool0}
		\caption{After first pooling.}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{pool1}
		\caption{After second pooling.}
	\end{subfigure}
	\begin{subfigure}{.3\textwidth}
		\centering
		\includegraphics[width=\linewidth]{fc0}
		\caption{After max-pooling.}
	\end{subfigure}
	\caption{Computation in the neural network, showing intermediate values involved in the network (correctly) predicting that an input formula from the test set is unsatisfiable. Vectors of floating-point numbers are shown as grids.}
	\label{figure:computation}
\end{figure}

\subsection{Evaluation}
The network was evaluated on the balanced test set of 200 examples, as described.
Various metrics for accuracy are shown in Table \ref{table:network-evaluation}.
While these results are very promising, it should be emphasised that it is unclear how effective a train/test split is in this setting (since similar subgoals may occur in both sets, even with proper data hygiene), and that this network is not attempting to determine the satisfiability of arbitrary formulae, merely those that occur in proof attempts on the \mizarlarge{} dataset.

The higher precision and lower recall values are likely an artefact of the augmentation process.
However, even with these caveats, the network performance is surpising and is practically useful for improving proof search in this dataset.

@@

\subsection{Benchmark Results}
Table \ref{table:m2k-results} shows the total number of theorems proved using various configurations of \z3 and LeRNA on the M2k dataset.
\z3 ran for a full 10 seconds to establish baseline performance, then as an oracle for 20 milliseconds to determine the number of ``trivial'' problems.
LeRNA ran on an identical dataset, first without guidance from the neural heuristic, then with guidance.

>>>>

\documentclass[runningheads]{llncs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{mathpartir}
\usepackage{subcaption}
\usepackage{textcomp}
\newcommand{\lerna}{\textsc{Lerna}}
\newcommand{\z}[1]{\textsc{Z3}}
\newcommand{\mizarlarge}{\textit{M40k}}
\newcommand{\mizarsmall}{\textit{M2k}}
\begin{document}
\title{A Neural, Parallel Theorem Prover*}
\author{
	Michael Rawson*\orcidID{0000-0001-7834-1567}\inst{1} \and
	Giles Reger\inst{2}
}
\institute{
	University of Manchester, UK\\
	\email{michael@rawsons.uk}\and
	University of Manchester, UK\\
	\email{regerg@cs.man.ac.uk}
}
\maketitle

@@

\section{Introduction}
Recent advances~ in neural network systems allow for processing graph-structured data in a neural context.
Graphs are a natural representation for logical formulae as found in automatic theorem provers (ATPs), suggesting a new breed of \emph{neural ATP} in which proof search is guided by a neural black-box acting as ``mathematician's intuition''.
However, in practice there are several implementation issues which must be avoided in order for neural systems to integrate with efficient traditional ATPs:
\begin{enumerate}
	\item Proof state in such systems may be very large, such as in saturation-based provers, leading to training data which is impractical to learn from and slow to evaluate.
	\item Data structures employed may be very opaque or ``unnatural'', containing artifice designed for efficiency rather than natural comprehension by a neural network.
	\item Systems may be very sensitive to latency, which can result in the introduction of neural guidance systems crippling prover throughput and hence performance.
\end{enumerate}
%
Attempting to solve these issues with a novel prover architecture, and exploring several options to improve overall efficiency, the prototype system \lerna{}\footnote{\textbf{Le}arning to \textbf{R}eason with \textbf{N}eural \textbf{A}rchitectures. Lerna is also the lair of the mythical many-headed beast \textsc{Hydra}.} takes an alternative step toward useful neural automatic theorem provers.

\section{Background}
An understanding of neural networks~\cite{neural-survey}, and of first-order logic and automatic theorem provers~\cite{fol-atp-survey} is assumed.
The particular syntax of first-order logic used in this work is presented here as clarification.

\subsection{Syntax of First-Order Logic}
Terms \(t\) are either
\begin{enumerate}
	\item Variables \(x\), bound by a quantifier.
	\item Function applications \(f\left(t_1, t_2, \ldots, t_n\right)\) where \(f\) is an \(n\)-ary function symbol. Constants \(c\) are a degenerate case in which \(n = 0\).
\end{enumerate}
%
Formulae \(\phi, \psi\) are one of
\begin{enumerate}
	\item Truth \(\top\) or falsehood \(\bot\).
	\item Equalities \(t_1 = t_2\).
	\item Predicate applications \(p\left(t_1, t_2, \ldots, t_n\right)\), where \(p\) is an \(n\)-ary predicate symbol.
	\item Negations \(\lnot \phi\), implications \(\phi_1 \Rightarrow \phi_2\), or equivalences \(\phi \equiv \psi\).
	\item Conjunctions \(\phi_1 \wedge \phi_2 \wedge \ldots \wedge \phi_n\) or disjunctions \(\phi_1 \vee \phi_2 \vee \ldots \vee \phi_n\).
	\item Universal or existential quantifiers \(\forall x. \phi, \exists x. \phi\).
\end{enumerate}

\subsection{Modern Automatic Theorem Provers}
State-of-the-art ATP systems (e.g. E~\cite{E}, iProver~\cite{iProver}, Vampire~\cite{Vampire}, CVC4~\cite{CVC4}) utilise efficient proof calculi combined with highly-configurable search routines in order to explore a large search space efficiently.
Generally, the search space in these systems slowly at first, then increases rapidly, leading to the maxim that a proof is typically found quickly or not at all.
Portfolio modes~\cite{portfolio} are therefore employed in order to achieve good performance on a large set of practically-useful problems, such as those found in TPTP~\cite{TPTP}.

\subsection{Machine Learning and Theorem Proving}
Despite the efficiency of such systems, they can still spend time exploring areas that a human mathematician would discard, and tuning such systems manually is in general extremely difficult~\cite{portfolio}.
There is therefore interest in applying machine-learning techniques to automatic theorem provers, with the eventual aim of an ``intelligent'' theorem prover which can learn from past experience to develop an intuition, discard uninteresting search space, and tune performance in a more principled way.

Work has proceeded in the fields of premise selection~\cite{DeepMath,MaSh,MaLARea,formula-graph}, static strategy selection~\cite{Bridge,MaLeS,E-MaLeS}, dynamic (run-time) strategy selection~\cite{dynamic-strategy-priority} and more recently, direct proof guidance~\cite{MaLeCoP,FEMaLeCoP,rlCoP,ENIGMA,DNGPS}.
Proof guidance typically involves some form of machine-learned \emph{heuristic} which biases proof search in some way, allowing the prover to avoid parts of the search space deemed uninteresting by the heuristic.

Much previous work on integrating machine-learned heuristics into automatic theorem provers has relied on hand-engineered features~\cite{MaLeCoP,FEMaLeCoP,rlCoP} or other embedding methods~\cite{ENIGMA}, which have the advantage of simplicity and relative efficiency, but do not encode all information available.
By contrast, a neural method which takes into account all information should allow for greater precision in proof guidance systems.
Deep Network Guided Proof Search (DNGPS)~\cite{DNGPS} is an example of previous work in this area, which integrated a deep neural guidance system into the saturation-based prover E~\cite{E}.
DNGPS achieved successful results, but suffered from the latency introduced into the system by the neural heuristic: despite processing only a reduced amount of the available proof state, the reduction in throughput necessitated a two-phase approach in which the prover was neurally-guided in the first phase, before falling back to traditional proof search in the second.
rlCoP~\cite{rlCoP} is a connection-based reinforcement-learning prover which is not presently neurally-guided, but takes a similar approach to that taken in this paper and achieves impressive results.

\subsection{Neural Networks for Formulae}
Neural networks are well-known tools for supervised learning~\cite{neural-survey}, and combined with trainable convolution/pooling operators are suitable for processing large-scale data such as images~\cite{cnn}.

Processing structured data such as logical formulae is a relatively new domain for neural networks.
Some work attempts to use unstructured representations of such formulae, such as text, or build entirely-new models for a specific logic~\cite{logical-entailment}, whereas others attempt to re-use neural techniques for generic structures such as trees~\cite{logical-tree}.
A promising direction in this area is recent research on neural methods working with graphs~\cite{graph-cnn,gcn,gcn-relational}, which have already been applied to premise selection~\cite{formula-graph}.
Graph neural networks tend to include network layers inspired by convolution operators in image-processing networks, combining information from neighbouring nodes (pixels).
For example, the Graph Convolutional Network~\cite{gcn} includes the following convolution layer, where \(H_i\) is the network output at layer \(i\), \(\bar{A}\) is the self-connected adjacency matrix, \(\bar{D}\) is the diagonal node degree matrix of \(\bar{A}\), \(W_i\) is a trainable weight matrix and \(\sigma\) is a non-linear activation function:
\[
	H_{i + 1} = \sigma\left(\bar{D}^{-\frac{1}{2}}\bar{A}\bar{D}^{-\frac{1}{2}}H_iW_i\right)
\]

@@

\section{Design}
\label{section:Design}
In order to achieve the goal of a neural theorem prover without the disadvantages associated with neural approaches, a new design of theorem prover is required.
Popular calculi used in existing ATPs tend to be unfriendly to neural guidance.
For such a system, we require the following from the calculus:
\begin{enumerate}
        \item \emph{Proof state must be small.} Attempting to evaluate large proof states structurally requires a lot of computation and resources. Saturation-based provers can have very large proof states, for example.
        \item \emph{Evaluation of states must be possible in parallel.} Machine-learning algorithms tend to operate more efficiently in batches. Tree-based approaches (tableau \emph{etc.}) lend themselves to this, whereas saturation provers are inherently sequential.
        \item \emph{Subgoals must be independent and self-contained.} If the prover has a notion of (sub-)goals which must be dispatched (such as in tableau provers), these should be independent of the rest of the search space, without e.g. unifiers. Otherwise, the learning system is trying to learn while blind to the context of the search.
	\item \emph{Subgoals must be intelligible.} Adding ``noise'' such as clausification obscures the original intuition behind a goal, at least for human observers. While this is not necessarily the case for machine-learning algorithms, it seems likely that removing structure and adding artefacts will reduce model performance.
\end{enumerate}
We therefore implement a refutation prover based on a first-order tableaux calculus without unification, on non-clausal formulae.
Each goal in this case is the set of formulae present on the tableau branch.
In this context, proof state is small (only the current branch), evaluation of states is possible in parallel, each branch is independent and contains all information required, and all available structure from the original problem is kept.

\subsection{Search}
In the calculus (see Section \ref{section:Calculus}) for this prover, there are two branching factors: each goal has a set of possible inferences, and each inference contains a set of possible sub-goals.
To prove a goal, at least one inference must be proved.
To prove an inference, all the inferences' sub-goals must be proved.
A simple optimisation observation is that sub-goals might well be shared between inferences from all over the search space, so search becomes a directed acyclic graph, alternating between goals and inferences.

\begin{figure}
	\centering
	\includegraphics[width=0.6\linewidth]{search}
	\caption{Search in the \lerna{} system, showing shared sub-goals.}
\end{figure}

Now the search graph can be explored: in each step, a leaf (goal) node is selected for expansion, and all resulting inferences and sub-goals are added to the graph.
If a goal has no possible inferences, it is satisfiable and can be removed from the search space.
On the other hand, if a goal is trivial (i.e. contains a contradiction), it is unsatisfiable and can be marked as proven.
This idea can be lifted to inferences: if an inference contains any satisfiable sub-goal, it too is satisfiable, whereas if an inference contains all unsatisfiable goals, it is unsatisfiable.
Proof search continues until the timeout is reached or the root goal is shown to be (un-)satisfiable.
In order to dispatch trivial sub-goals quickly, an existing fast \emph{oracle} ATP is used (see Section \ref{section:oracle}).
This may mark goals as (un-)satisfiable, at which point no further exploration is required.

Search is biased by heuristic evaluation.
The neural heuristic function (see Section \ref{section:heuristic}) evaluates each goal and assigns a score corresponding to whether the network believes that the goal is satisfiable or unsatisfiable.
In order to balance exploitation of promising directions and exploration of all parts of the search space, a principled UCT-based search algorithm is used, as in MonteCoP~\cite{MonteCoP}.
At each sub-goal, the prover chooses the inference maximising the expression
\[
	s + c\sqrt{\frac{\ln N}{n}}
\]
where \(s\) is the score associated with the inference, \(c\) is the exploration parameter (theoretically \(\sqrt{2}\)), \(N\) is the number of visits to the current node, and \(n\) is the number of visits to the child.
After an inference is selected, the least-promising sub-goal is selected, and the selection process continues until a leaf node is reached.
The score \(s\) of a leaf node is the score assigned to it by the heuristic function.
An inference's score is the minimum of their sub-goals' scores, and non-leaf sub-goal's score is the maximum of their inference's scores.

\subsection{Architecture}
The system aims to consume all available CPU and GPU resources as efficiently as possible.
To that end, proof search is asynchronous: the search algorithm generates new sub-goals, which are placed on two separate queues: one for the oracle ATP, another for heuristic evaluation.
Proof search then continues elsewhere, while the oracle ATP is called in parallel on each sub-goal (consuming all available CPU) while the heuristic consumes batches of subgoals, efficiently utilising the available GPU resource.
As information flows backwards from these processes, the search process updates its information about a given sub-goal and propagates that information upwards to the sub-goal's parent inferences, to influence future proof search: see Figure \ref{figure:architecture}.

\begin{figure}
	\includegraphics[width=\linewidth]{architecture}
	\caption{A cartoon showing information flow in the system.}
	\label{figure:architecture}
\end{figure}

\subsection{Implementation}
The Rust~\cite{rust} programming language was chosen to implement the main prover (minus the heuristic), for its memory and thread-safety guarantees, as well as its support for managing memory allocation and algebraic datatypes.
The implementation is currently just under 3,000 lines of Rust, not including \z3{}, the TPTP format parser or the implementation of perfect sharing.
Python 3~\cite{python3} was used for the heuristic due to the large number of libraries available for neural network implementation in Python.
The heuristic is implemented as a server, communicating with the main prover via a TCP socket.
In principle this allows for the heuristic to be a shared resource with a centralised heuristic server, or a load-balanced cluster.

\section{Calculus}
\label{section:Calculus}
The proof calculus used in this context may be extremely general: in fact, any function from goals to a finite set of possible inferences (themselves finite sets of sub-goals) will suffice, as long as each goal remains independent of any other such that the heuristic function can process all available information.  If the inference system is complete, there are no additional constraints such as orderings or fairness to ensure the completeness of the prover, as the balanced search algorithm (see Section \ref{section:Design}) will ensure this.

\lerna{} presently implements a refutation tableaux calculus~\cite{handbook-of-automated-reasoning} without unification.
The calculus described is deliberately na\"ive in order to easily satisfy the design constraints described, but this is a weak point of this work and will be improved in future.
A na\"ive calculus is not necessarily a problem in this context, as the heuristic should in principle select promising areas to explore and ignore uninteresting sub-goals.
However, a more efficient calculus would improve performance where the heuristic fails.

\subsection{Refutation Tableaux}
In order to show a conjecture \(C\) from a set of axioms \(A_i\), it suffices to negate \(C\) and the show that the resulting conjunction \(A_1 \wedge A_2 \wedge \ldots \wedge \lnot C\) is unsatisfiable.
A set of inference rules of the form
\begin{mathpar}
	\inferrule{\Gamma_1 \\ \Gamma_2 \\ \ldots \\ \Gamma_n}{\Delta}
\end{mathpar}
where \(\Gamma_i, \Delta\) are sets of formulae and \(\lnot\left(\Gamma_1 \wedge \Gamma_2 \wedge \ldots \Gamma_n\right) \Rightarrow \lnot \Delta\) is an unconditional tautology, form a refutation calculus.
Proofs in this calculus can be expressed by closed trees of inference rules.

\subsection{Complete Inferences}
The inference rules in Figure \ref{figure:complete-inferences} form a complete inference system, by analogy with a first-order tableaux calculus without unification.
A difference and point of interest is the rule for instantiating universal quantifiers: instead of instantiating a variable with any possible term \(t\) --- an infinite space --- it is instantiated with one function symbol (or constant) at a time, quantifying over new variables as needed.
This allows for instantiating any term over multiple inference rules (effectively enumerating the Herbrand universe for the goal), but without an infinite number of possible inferences at any point.
Equality is handled by a rule rewriting classes of equal ground terms.
Both of these rules are complete yet inefficient, but both are likely to be used only a few times in order to provide enough of a ``hint'' to the oracle system for it to find a proof.

\begin{figure}
	\centering
	\begin{mathpar}
		\inferrule[contradiction]{ }{\phi, \lnot \phi, \Gamma}\and
		\inferrule[equal]{t = s, \phi\left[t/s\right], \Gamma}{t = s, \phi, \Gamma}\and
		\inferrule[implies]{\lnot \phi, \psi, \Gamma}{\phi \Rightarrow \psi, \Gamma}\and
		\inferrule[equivalent]{\lnot \phi, \lnot \psi, \Gamma\\\phi, \psi, \Gamma}{\phi \equiv \psi, \Gamma}\and
		\inferrule[conjunction]{\phi_1, \phi_2, \ldots, \phi_n, \Gamma}{\phi_1 \wedge \phi_2 \wedge \ldots \wedge \phi_n, \Gamma}\and
		\inferrule[disjunction]{\phi_1, \Gamma \\ \phi_2, \Gamma \\ \ldots \\ \phi_n, \Gamma}{\phi_1 \vee \phi_2 \vee \ldots \vee \phi_n, \Gamma}\and
		\inferrule*[lab={instantiation}]{\forall x_1, x_2, \ldots x_n. \phi[f(x_1, x_2, \ldots x_n)/x], \forall x. \phi, \Gamma}{\forall x. \phi, \Gamma}\and
		\inferrule[non-empty]{\phi[k/x], \Gamma}{\forall x. \phi, \Gamma}\and
		\inferrule[exists]{\phi[k/x], \Gamma}{\exists x. \phi, \Gamma}
	\end{mathpar}
	\caption{A complete inference system for \lerna{}. Rules for negation are as usual and not shown here for brevity. In rule \textsc{instantiation}, \(f\) is a function symbol of arity \(n\) in the conclusion's signature and \(x_1 \ldots x_n\) are fresh for the conclusion. In rules \textsc{non-empty} and \textsc{exists}, \(k\) is fresh for the conclusion. \(\phi[t/s]\) is a capture-avoiding substitution replacing \(t\) for \(s\) in \(\phi\).}
	\label{figure:complete-inferences}
\end{figure}

\subsection{Weakening}
A weakening rule is an important part of \lerna{}'s calculus, since the \textsc{instantiation} and \textsc{equal} rule can produce a large number of formulae, some of which must be removed to help the oracle to prove the goal.
Each application of the rule removes some amount of information from the goal in order to simplify it --- this is sound and corresponds to removing an axiom from proof search.
The rule is merely
\begin{mathpar}
	\inferrule[weaken]{\Gamma}{\phi, \Gamma}
\end{mathpar}

\subsection{Simplifications}
Before each inferred goal is added to proof search, it is simplified with a simplification function mapping a goal to a possibly-simpler goal.
This allows for efficient removal of tedious inferences such as double-negation elimination and generally reduces the search space.
A set of example simplification rules are shown in figure \ref{figure:simplification-rules}.

\begin{figure}
	\centering
	\begin{mathpar}
		\inferrule[double-negation]{\phi, \Gamma}{\lnot \lnot \phi, \Gamma}\and
		\inferrule[conjunction-associative]{\phi \wedge \psi \wedge \pi, \Gamma}{\phi \wedge \left(\psi \wedge \pi\right)}\and
		\inferrule[disjunction-propagation]{\phi, \Gamma}{\phi \vee \bot, \Gamma}\and
		\inferrule[reflexivity]{\top, \Gamma}{t = t, \Gamma}\and
		\inferrule[free]{\phi, \Gamma}{\forall x. \phi, \Gamma}
	\end{mathpar}
	\caption{Some simplification rules implemented in \lerna{}. In rule \textsc{free}, \(x\) is free in \(\phi\). Several other rules are implemented.}
	\label{figure:simplification-rules}
\end{figure}

@@

\subsection{Acting as a Preprocessor}
\lerna{} as a whole might also be seen as an intelligent preprocessor for existing ATPs in this setting: existing theorem provers are known to be sensitive to small changes in their input~\cite{clausification}, and generally make little attempt to split their input into smaller sub-goals, for parallelism~\cite{parallel} or otherwise.
The system can therefore act as an adaptor for any existing ATP, adding parallelism opportunities and ``smoothing out'' sensitivity to input syntax.

\section{Learned Heuristic}
\label{section:heuristic}
A suitable heuristic function for the system must predict a value between 0 and 1 for a given formula \(F\), where 0 represents a satisfiable goal and 1 represents unsatsfiability, based on a set of tagged formulae seen in previous proof search.
Although the data is collected by running the system itself (and therefore might be considered \emph{reinforcement} learning~\cite{reinforcement}), for this approach data collection and learning were considered separately and hence forms a classic supervised-learning problem.

@@

\begin{figure}
	\centering
	\begin{subfigure}{.35\textwidth}
		\centering
		\includegraphics[width=\linewidth]{translation-ast}
		\caption{AST.}
	\end{subfigure}
%	\begin{subfigure}{.39\textwidth}
%%		\centering
%		\includegraphics[width=\linewidth]{translation-graph}
%		\caption{De-duplicated DAG.}
%	\end{subfigure}
	\begin{subfigure}{.3\textwidth}
		\centering
		\includegraphics[width=\linewidth]{translation-nameless}
		\caption{Nameless DAG.}
	\end{subfigure}
	\begin{subfigure}{.25\textwidth}
		\centering
		\includegraphics[width=\linewidth]{translation-undirected}
		\caption{Final result.}
	\end{subfigure}
	\caption{The translation process for \(\forall x. \left[p\left(f(c), x\right) \wedge \lnot\left(f(c) = d\right)\right]\) to a graph, as seen by the neural network.}
	\label{figure:translation}
\end{figure}

\subsection{Augmentation}
\label{section:augmentation}
One possible solution~\cite{imbalanced} to the problem of classification on imbalanced domains is to synthesise new data for under-represented classes --- in this case unsatisfiable formulae --- from existing data by augmenting it.
An example is augmenting image data by cropping, flipping or adding noise to existing images.
There are many possible ways to augment formulae graphs.
For this prototype, a simple approach is taken in which a small number of nonsense formulae are added to the graph by randomly adding nodes/edges where appropriate.
This approach has the advantage of exposing the network to ``noise'' such as additional axioms which might well occur in practice, but if the network is adequately capable of filtering these then no new formulae are actually seen.

\subsection{Neural Architecture}
In a typical convolutional network architecture for images~\cite{cnn}, there are a series of filtering stages, followed by a densely-connected neural network.
Each filtering stage intuitively combines data from local features (via \emph{convolution}), then reduces the dimensions of the image (via \emph{pooling}) for the next stage.
Graph neural networks have analogous convolution~\cite{gcn} (combining information from neighbouring nodes) and pooling~\cite{top-k-pooling} (merging nodes to reduce the size of the input graph) operators.
A brief period of experimentation with these operators yielded the following network architecture, shown in Figure \ref{figure:network}.
It is not claimed that this is the optimal configuration, and no grid search has yet taken place to optimise the network parameters.
\begin{enumerate}
	\item Input. a graph \(G\) consisting of one-hot encoded nodes \(N\) and edges \(E\).
	\item Embedding. Each node is mapped to an embedding vector of size 64 via a trained dense embedding.
	\item Initial Convolution. 4 convolution layers are applied to the graph with rectified linear activations. This yields a graph of the same size, but with information exchanged between nodes.
	\item Convolution/Pooling. Similar convolution layers are then passed through \emph{top-\(k\)}~\cite{top-k-pooling} layers, retaining \(k = 60\%\) of the graph's nodes. This is repeated 3 times, reducing the size of the graph considerably.
	\item Convolution/Max-Pooling. A final convolution layer feeds into a max-pooling layer, combining all remaining node data into one datum, and dropping the edge data.
	\item Fully Connected. A fully-connected hidden layer with rectified linear activation halves the input size.
	\item Fully Connected/Softmax. A fully-connected final layer outputs two class labels, with softmax activation.
\end{enumerate}

@@

\subsection{Implementation and Training}
This architecture was implemented with the PyTorch~\cite{pytorch} neural network library, combined with a graph-processing (``geometric'') extension library, PyTorch Geometric~\cite{pytorch-geometric}, which together provide facilities for automatic differentiation, GPU-accelerated training, pre-programmed layers for graph processing, and various utilities.
The dataset is split into a large training set and a smaller test set (200 balanced examples), since unsatisfiable examples were time-consuming to obtain in this setting.
The unsatisfiable training data were then augmented as described in section \ref{section:augmentation} to produce a balanced total training set of 3.5 million examples.
The network was trained on commodity desktop hardware with a mid-range GPU~\footnote{NVIDIA\textsuperscript{\textregistered} GeForce\textsuperscript{\textregistered} GT 730.} for 8 epochs/24 hours, optimising a negative log-likelihood loss function.

\begin{figure}
	\centering
	\captionsetup[subfigure]{justification=centering}
	\begin{subfigure}{.3\textwidth}
		\centering
		\includegraphics[width=\linewidth]{embedding}
		\caption{Output of embedding layer.}
	\end{subfigure}
	\begin{subfigure}{.3\textwidth}
		\centering
		\includegraphics[width=\linewidth]{conv}
		\caption{Output after initial convolutions.}
	\end{subfigure}
	\begin{subfigure}{.3\textwidth}
		\centering
		\includegraphics[width=\linewidth]{pool0}
		\caption{After first pooling.}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=\linewidth]{pool1}
		\caption{After second pooling.}
	\end{subfigure}
	\begin{subfigure}{.3\textwidth}
		\centering
		\includegraphics[width=\linewidth]{fc0}
		\caption{After max-pooling.}
	\end{subfigure}
	\caption{Computation in the neural network, showing intermediate values involved in the network (correctly) predicting the satisfiability of an input formula.}
	\label{figure:computation}
\end{figure}

\subsection{Evaluation}
The network was evaluated on the balanced test set of 200 examples, as described.
Various metrics for accuracy are shown in Table \ref{table:network-evaluation}.
While these results are very promising, it should be emphasised that it is unclear how effective a train/test split is in this setting (since similar subgoals may occur in both sets, even with proper data hygiene), and that this network is not attempting to determine the satisfiability of arbitrary formulae, merely those that occur in proof attempts on the \mizarlarge{} dataset.
The higher precision and lower recall values are likely an artefact of the augmentation process.
However, even with these caveats, the network performance is surprising and is practically useful for improving proof search in this dataset.

@@

\subsection{Benchmark Results}
Table \ref{table:m2k-results} shows the total number of theorems proved using various configurations of \z3 and \lerna{} on the \mizarsmall{} dataset.
\z3 ran for a full 10 seconds to establish baseline performance, then as an oracle for 20 milliseconds to determine the number of ``trivial'' problems.
\lerna{} ran on an identical dataset, first without guidance from the neural heuristic, then with guidance.
